---
title: "Assignment 1"
author: "Walter Verwer         Bas Machielsen"
date: "1/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stargazer)
library(readr)

```

## Question 1

1. The sample selection model. A researcher aims to gain insight in the potential earnings of the non-employed. (In the data, the non-employed can be identified by a missing value for the earnings variable). She realizes that the sample of observed wages may be subject to sample selection.

(a) Run an OLS regression for log-earnings on schooling, age, and age squared. Present the results and comment on the estimates.
    
```{r, include=FALSE}
data <- read_csv("./Data/logEarnings.csv")
```
    
```{r, results='asis'}
model <- lm(data = data, formula = logWage ~ schooling + age + age2)
stargazer(model, type = "html")
```

The results show that 1 additional year of schooling has an effect of 0.216 on log(Wage), which means that it has an effect of `r exp(0.216)` on schooling. Hence, each year of schooling supposedly increases earnings with `r exp(0.216)`. 
