---
title: "Assignment 5"
author: "Bas Machielsen & Walter Verwer"
date: \today
output:
  pdf_document:
    includes:
      in_header: "preamble.tex"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Register an inline hook (set rounding to 3 digitis):
# knitr::knit_hooks$set(inline = function(x) {
#   x <- sprintf("%1.3f", x)
#   paste(x, collapse = ", ")
# })

library(stargazer); library(tidyverse); 
library(sandwich);library(haven)
library(knitr); library(lmtest); library(sandwich); library(ivreg)

# Import Stata dataset
data <- haven::read_dta("./data/marriagemarket.dta")
attach(data)
```

## Problem 1

We assume the following regression model: $Y_{t,g} = \alpha_{t,g} + \delta \cdot D_{t,g} + \eta_{g} + u_{t,g}$ for $g \in \{T,C\}$ and $t \in \{0,1\}$. Then, the difference-in-difference estimator equals $\hat{\delta} = \delta \cdot D_1 + \alpha_{1,T} - \alpha_{0,T} + U_{1,T} - U_{0,T} - [\alpha_{1,C} - \alpha_{0,C} + U_{1,C} - U_{0,C}]$. The ATT = expected value of the DiD estimator is then:

\[
\mathbb{E}[\hat{\delta}|D=1] = \delta + [\alpha_{1,T} - \alpha_{0,T}] - [\alpha_{1,C} - \alpha_{0,C}]
\]

Hence, the expected value of $\hat{delta}$ depends on the assumption that the sum of the terms containing the $\alpha$'s equal zero, in other words, if there is a common time trend between treatment and control groups. 

If we assume that the program is known beforehand (by the students), and grades are a function of effort and ability $\in \{ \text{High, Low} \}$, and high-ability students are all in the treatment group, then a fraction of the treatment group will also consist of high-effort and low-ability students, whereas the control group will consist of low-ability students only. After being provided with the incentive of housing, they will readjust their effort in the 2nd year, and hence, obtain lower grades. This causes a violation of the common trend, because the low-ability individuals who are in the treatment group will revert back to their effort level that is unincentivized by housing. 

If we assume the program is not known beforehand, the students have no differing incentives, irrespective of their ability and effort. Hence, the common time trend assumption is justified and the estimator is unbiased. 

## Problem 2

__(i) Regress the number of out-of-wedlock births on the sex ratio, using only the observations from the pre-war period. Discuss your result. How can a difference-in-differences approach using the military mortality rate during WWI improve on this estimation strategy?__

```{r frederik, results='asis'}
model1 <- data %>%
    filter(post == 0) %>%
    lm(formula = "illeg ~ sr")

cov1 <- vcovHC(model1, type = "HC1")
robust_se <- sqrt(diag(cov1))

stargazer(model1, 
          dep.var.labels = "Illegal Births",
          omit.stat = "ser",
          header = F,
          se = list(robust_se))
```

The estimates focus on the cross-section and relies on between-department variation to find the correlation between the sex-rate and illegal births. This gives a biased estimate, because the (unaccounted for) department effects might be correlated with the sex ratio. The difference in difference estimator can improve on this by considering the within-departement estimate, thereby eliminating time- and department-specific effects from the estimate for the influence of the sex ratio. 

__(ii) Generate a dummy variable that indicates wether the military mortality in a region is above the median military mortality or not. Make a table with the mean percentage of out-of-wedlock births for the high and low mortality regions, both before and after the war. Use the numbers from the table to calculate the difference-in-differences estimator.__

``` {r q2, message = FALSE}
data <- data %>%
    mutate(htm_mortality = if_else(mortality > median(mortality, na.rm = T), 
                                   1, 
                                   0)
           )

table <-  data %>%
    filter(!is.na(htm_mortality)) %>%
    group_by(post, htm_mortality) %>%
    summarize(mean_illeg = mean(illeg, na.rm = T)) 

kable(table)

did <- (6.154916 - 5.086213) - (8.448658 - 7.960866)

```

The difference in difference estimator is equal to `r did`, indicating an increase of illegal births by `r did` percentage points, consistent with the hypothesis. 

\clearpage 

__(iii) Estimate the following model, which estimates the difference-in-differences estimator in a regression equation. What is the interpretation of the coefficients $\beta_1$ and $\beta_2$? What do you conclude about the effect of male scarcity on the number of out-of-wedlock births?__

```{r q3, echo = FALSE, results='asis'}
model_q3 <- lm(data = data, 
               formula = "illeg ~ post + post:mortality")

cov1 <- vcovHC(model1, type = "HC1")
robust_se <- sqrt(diag(cov1))

stargazer(model_q3, 
          dep.var.labels = "Illegal Births",
          omit.stat = "ser",
          header = F,
          se = list(robust_se))

```

The coefficient $\beta_1$ represents the average illegal births rate post-war, and the constant term represents the average illegant births pre-war. The coefficient $\beta_2$ represents the effect on illegal births of a 1-percentage point increase in mortality. The coefficient is negative, but not significant. The point estimate is, however, in accordance with the hypothesis, in that an increased mortality leads to an increase in out-of-wedlock births. 

The coefficient, however, might be biased because the model does not take into account systematic differences in the out-of-wedlock birth rate between various departments. 





